Fast Sorting on Modern Computers

### 2005-06-16 docs/tech_report/merge.tex
wrote up the software predictor section

\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_tiled_mergesort_0.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_0.eps}{mergesorts}
%\afterpage{
%\setcounter{subfigure}{0}
%\thispagestyle{empty}
%\clearpage
%\enlargethispage{14em}
%\vspace*{-5em}
%\begin{figure}[H]
%\begin{changemargin}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_n-0.eps}}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/algorithm_s-0.eps}}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/base_mergesort_0.eps}}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_tiled_mergesort_0.eps}}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort_0.eps}}
%\subfigure[]{\includegraphics[scale=0.74]{../counters/double_multi_mergesort-1.eps}}
%\end{changemargin}
%\vspace*{6em}
%\caption{Branch prediction performance for mergesorts. All graphs use the same scale.}
%\label{Branch prediction performance for mergesorts}
%\end{figure}
%\newpage
%}

%TODO This is a bit to rewrite

figure a shows Algorithm N, knuths natural merge. N3, N5 and N9 indicate parts of the flow chart. 'N3 i==j' is the second branch in N3.

The results show that the secondary loops are always taken. Taken here means broken out of, hence seldom used. The main action occurs in N3, the main comparator, which has a 50\% prediction ratio.

Algorithm S in graph B shows the same. I never really understood how it worked though.

Figure c shows base mergesort. This shows a lower number of branches and a higher percentage of correct predictions. The lower number of branches is probably due to the presort. there is a higher percentage of correct predictions, but the actual number of correct predictions in the same. This implies that doing the presort removes incorrect predictions (or, more to the point, moves incorrect predictions to another part of the algorithm). Note too, that both sides are used equally, which doesn't happen later. Two thirds of the branches are not-taken, which means loops execute twice before breaking. this is the same results that we found in quicksort later. I'm sure knuth explains this somewhere in the fine print.

Figure d shows double tiled mergesort. The total number fo branches is a 10\% increase over the base mergesort. Not sure why this is. 15\% increase in correct predictions so this incrase into predictable range, which should be avoidable (according to what we guess in quicksort).

Figure e and f show the double mulitmergesort results. This sorts fully within the cache only, reducing the number of branches relating to this. The reason that the previous balance was so muhc in favour of forwards was that the final sort at the end was forwards. Here though, the arrays are left bitonically, so there is an even distribution betwwen forwards and backwards. There si still the bias towards left rather than right. Left comes first in the code, but this doesnt quite explain why this happens. However, the number fo misses each way is identical.

The total branches is a decrease of about 20\% on base mergesort. The number of mispredictions decreases by less than 10\%. Howver, all these branches are made up by the increase in branches due to the heap. The number of extra brnahc misses due to this is 26\% of the branches misses in base mergesort. In total, the num ber fo branches isa equivalent however. So no loss by doing this (except lots of extra instructions, as seen before).

Note that here and in the heaps, comparison 0 does nothing. THe compiler ought to trim it out, except that now it actually does something (increases the branch counter).

finally, in base mergesort, there is an inline presort used. For efficiency, an insertion sort is used later. However, there are 6.8 misses per key in the inline presort, which is less than the insertion sort does. Its not as useful for presorting to 8, however. THese results hold forwards and in reverse.

%TODO up to here

@@

%TODO
TODO what about level 1 cache misses if we realigned by 1024?

>>>>

\sixplot{algorithm_n_0.eps}{algorithm_s_0.eps}{base_mergesort_0.eps}{double_tiled_mergesort_0.eps}{double_multi_mergesort_0.eps}{double_multi_mergesort_0.eps}{mergesorts}

Figure \ref{mergesorts-1} shows that algorithm N has the same number of branches
and branch mispredictions as algorithm S, shown in \ref{mergesorts-2}. Base
mergesort has the same number of mispredictions (see Figure \ref{mergesorts-3},
indicating a possible minimum number. It has a lower number of branches,
probably due to performing a persort first. The presort has approximately 6.8
misses per key (not shown in the graphs). In later mergesorts, an insertion sort
is used instead.
%TODO ohw much better does this do? if not, move to other section

Double-tiled mergesort (which has the same results are tiled mergesort) has a
predilication for left and for merging forwards. This is likely to be due to the
fact that the final merge goes forwards, and that left occurs first in the code.

In both double tiled mergesort and base mergesort, approximately two thirds are
of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
an increase in branches, as opposed to a decrease in mispredictions. Overall
this stay is line with the idea that there is a minimum number of branch misses,
below which comparative sorts will not drop.

Double multi-mergesort sorts the arrays fully in the cache, leaving them
bitonically arranged. This means that there is no bias for the forward branches,
as there was in tiled mergesort. This is shown in Figures \ref{mergesorts-5} and
\ref{mergesorts-6}. While merging the number of branches is 20\% lower than base
mergesort, while the number of mispredictions in only 10\% lower. This implies
that most of the mispredictions occur early in the merging process, and not in
the final merge, which multi-mergesort leaves out.

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversly affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{cache heapsort
curve}). Note that \n{comparison 0} occurs every time, and would be optimized
out by the compiler.

@@
